# Production of HYSPLIT trajectory files {#prod-hyspl-traj}

As discussed in Section \@ref(trajPlot), [openair]{.pkg} can import
pre-calculated trajectory data for specified locations. The data are
stored on a Ricardo webserver to make it easy to import
96-hour back trajectory data. Several users have requested how they
can run HYSPLIT themselves e.g. for different trajectory start heights
or for many locations. This section provides the code necessary to run
the HYSPLIT model. The code below assumes that full years are run, but
it could be adopted for shorter periods. There are three main parts to
producing trajectory files:

1.  Download and install the NOAA Hysplit model, somewhere with
  write access (see below).

1.  Download the monthly meteorological (.gbl) files also from the
  NOAA website.

1.  Obtain the code to run Hysplit.


To run back trajectories it is necessary to download the
meteorological data files. The easiest way to download the
meteorological files is using the function below.

```{r getMet, eval=FALSE}
getMet <- function (year = 2013, month = 1, 
                    path_met = "~/TrajData/") {

    for (i in seq_along(year)) {
        for (j in seq_along(month)) {
            download.file(url = paste0("ftp://arlftp.arlhq.noaa.gov/archives/reanalysis/RP",
                              year[i], sprintf("%02d", month[j]), ".gbl"),
                          destfile = paste0(path_met, "RP", year[i],
                              sprintf("%02d", month[j]), ".gbl"), mode = "wb")
        }
    }
}
```

The function will download monthly met files (each about 120 MB) to
the chosen directory. Note that the met data files only need be
downloaded once. For example, to download files for 2013:

```{r getMet2, eval=FALSE}
getMet(year = 2013, month = 1:12)
```

The original functions have been modified by Stuart Grange (PhD
student at the University of York). It is first necessary on ensure
that the [stringr]{.pkg} and [devtools]{.pkg} packages are installed. The
latter is needed to load some R functions stored as a *GitHub
  gist* (some code that can be shared publicly).

```{r getGist}
library(devtools)
source_gist("https://gist.github.com/davidcarslaw/c67e33a04ff6e1be0cd7357796e4bdf5",
            filename = "run_hysplit.R")
``` 

Now there should be several loaded functions, including
`run_hysplit`. To run Hysplit, have a look at the examples here https://gist.github.com/davidcarslaw/c67e33a04ff6e1be0cd7357796e4bdf5.

On my Windows machine it is run as follows:

```{r run_hysplit, eval=FALSE}
data_out <- run_hysplit(
latitude = 36.134, 
 longitude = -5.347, 
  runtime = -96, 
  start_height = 10, 
  model_height = 10000, 
  start = 2015,
  end = "2015-01-10",
  hysplit_exec = "~/hysplit4/exec", 
  hysplit_input = "~/trajData", 
  hysplit_output = "~/temp",
  site = "gibraltar")
``` 

The `data_out` can then be used directly in [openair]{.pkg} trajectory
functions.

Most of the options should be self-explanatory but `hysplit_exec`
is the path to the Hysplit executable, `hysplit_input` is the path
to the meteorological files (downloaded as described above) and
`hysplit_output` is the directory where Hysplit will write its
temporary files.

Once run it is then advisable to store the data somewhere. Save it
like:

```{r saveHysplit, eval=FALSE}
saveRDS(data_out, file = "~/trajProc/myTrajData.rds")
``` 

Then it is easy to read in later and use e.g.

```{r importTrajLocal, eval=FALSE}
traj <- readRDS("~/trajProc/myTrajData.rds")
```
