--- 
title: "openair"
author: "David Carslaw"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib, DCC-refs.bib]
before_chapter_script: ["load-packages.R"]
always_allow_html: true
biblio-style: apalike
link-citations: yes
#github-repo: rstudio/bookdown-demo
description: "This will eventually form a book related to the openair R package."
---

# Prerequisites

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')

library(knitr)
library(tidyverse)
knit_hooks$set(crop = hook_pdfcrop)
opts_chunk$set(fig.align='center',fig.show='asis',warning=FALSE, crop=TRUE, background=c(1,1,1), cache=TRUE, tidy=FALSE)
```
