# The openair package {#sec:openair-package}

SECTION NEEDS UPDATING

In this book two packages are frequently used and it is a good idea
to load both.

```{r message=FALSE,warning=FALSE}
library(openair)
library(tidyverse)
``` 

Because the [openair]{.pkg} package (and R itself) are
continually updated, it will be useful to know this document was
produced using R version `r getRversion()` and [openair]{.pkg} version
`r packageDescription("openair", field = "Version")`.

## Access to source code
All R code is accessible. On CRAN, you will see there are various
versions of packages: Package source, MacOS binary and Windows
binary. The source code is contained in the Package source, which is a
tar.gz (compressed file). 

For [openair]{.pkg} all development is carried out using Github for version
control. Users can access all code used in openair at (https://github.com/davidcarslaw/openair).

## Input data requirements

The [openair]{.pkg} package applies certain constraints on input data
requirements. **It is important to adhere to these requirements
  to ensure that data are correctly formatted for use in [openair]{.pkg}.**
The principal reason for insisting on specific input data format is
that there will be less that can go wrong and it is easier to write
code for a more limited set of conditions.

- Data should be in a data frame (or `tibble`).

- **The date/time field should be called `date`** ---
  note the lower case. No other name is acceptable.

-  **The wind speed and wind direction should be named `ws` and
  `wd`**, respectively (note again, lower case). Wind directions follow the UK Met Office format and are represented as degrees from north
  e.g. 90 degrees is east. North is taken to be 360 degrees

-  Where fields should have numeric data e.g. concentrations of
  NO~x~, then the user should ensure that no other characters are
  present in the column, accept maybe something that represents
  missing data e.g. `no data'. .

-  Other variables names can be upper/lower case *but should
    not start with a number*. If column names do have white spaces, R
  will automatically replace them with a full-stop. While `PM2.5` as a
  field name is perfectly acceptable, it is a pain to type it
  in---better just to use `pm25` ([openair]{.pkg} will recognise pollutant names like this and automatically format them as PM~2.5~ in plots).

## Brief overview of `openair` {#sec:brief-intr-open}

This section gives a brief overview of the functions in [openair]{.pkg}. 
Having read some data into a data frame it is then straightforward to
run any function. Almost all functions are run as:

```{r eval=FALSE}
functionName(thedata, options, ...)
```

The usage is best illustrated through a specific example, in this case
the `polarPlot` function. The details of the function are shown in Section \@ref(sec:polarPlot) and through the help pages (type
?polarPlot). As it can be seen there are numerous options associated
with `polarPlot` --- and most other functions and each of these has
a default. For example, the default pollutant considered in
`polarPlot` is `nox`.  If the user has a data frame called
`theData` then `polarPlot` could minimally be called by:

```{r eval=FALSE}
polarPlot(theData)
```

which would plot a `nox` polar plot if `nox` was available in the data
frame `theData`.

Note that the options do not need to be specified in order nor is it
always necessary to write the whole word. For example, it is possible
to write:

```{r eval=FALSE}
polarPlot(theData, type = "year", poll = "so2")
```

In this case writing `poll` is sufficient to uniquely identify that
the option is `pollutant`.

Also there are many common options available in functions that are not
explicitly documented, but are part of lattice graphics. Some common
ones are summarised in Table \@ref(tab:options). The `layout`
option allows the user to control the layout of multi-panel plots
e.g. `layout = c(4, 1)` would ensure a four-panel plot is 4 columns
by 1 row.

```{r echo=FALSE}

tab_dat <- tibble(
  option = c("xlab", "ylab", "main", "pch", "cex", "lty", 
             "lwd", "layout"),
  description = c("x-axis label", "y-axis label", "title of the plot",
                  "plotting symbol used for points",
                  "size of symbol plotted",
                  "line type",
                  "line width",
                  "the plot layout e.g. c(2, 2) "
                  )
)
```

```{r options,echo=FALSE}
knitr::kable(
  tab_dat, booktabs = TRUE,
  caption = 'Common options used in [openair]{.pkg} plots that can be set by the user but are generally not explicitly documented.'
)
```

## The `type` option

One of the central themes in [openair]{.pkg} is the idea of
*conditioning*. Rather than plot $x$ against $y$,
considerably more information can usually be gained by considering a
third variable, $z$. In this case, $x$ is plotted against $y$ for many
different intervals of $z$. This idea can be further extended. For
example, a trend of NO~x~ against time can be *conditioned* in
many ways: NO~x~ vs. time split by wind sector, day of the week, wind
speed, temperature, hour of the day ... and so on. This type of
analysis is rarely carried out when analysing air pollution data, in
part because it is time consuming to do. However, thanks to the
capabilities of R and packages such as [lattice]{.pkg} and [ggplot2]{.pkg}, it becomes easier to work in this way.

In most [openair]{.pkg} functions conditioning is controlled using the
`type` option. `type` can be any other variable available in a
data frame (numeric, character or factor). A simple example of
`type` would be a variable representing a 'before' and 'after'
situation, say a variable called `period` i.e. the option `type
  = "period"` is supplied. In this case a plot or analysis would be
separately shown for 'before' and 'after'. When `type` is a numeric
variable then the data will be split into four *quantiles* and
labelled accordingly. Note however the user can set the quantile
intervals to other values using the option `n.levels`. For example,
the user could choose to plot a variable by different levels of
temperature. If `n.levels = 3` then the data could be split by
'low', 'medium' and 'high' temperatures, and so on. Some variables are
treated in a special way. For example if `type = "wd"` then the
data are split into 8 wind sectors (N, NE, E, ...) and plots are
organised by points of the compass.

There are a series of pre-defined values that `type` can take
related to the temporal components of the data. To use these there
*must* be a `date` field so that it can be calculated. These
pre-defined values of `type` are shown below are both useful and convenient. Given a
data frame containing several years of data it is easy to analyse the
data e.g. plot it, by year by supplying the option `type =
  "year"`. Other useful and straightforward values are "hour" and
"month". When `type = "season"` [openair]{.pkg} will split the data by
the four seasons (winter = Dec/Jan/Feb etc.). Note for southern
hemisphere users that the option `hemisphere = "southern"` can be
given. When `type = "daylight"` is used the data are split between
nighttime and daylight hours. In this case the user can also supply
the options `latitude` and `longitude` for their location (the
default is London).

And a brief summary of in-built types:

* **"year"** splits data by year
* **"month"** splits variables by month of the year
* **"week"** splits variables by week of the year
* **"monthyear"** splits data by year *and* month
* **"season"** splits variables by season. Note in this case the user can
  also supply a `hemisphere` option that can be either "northern"
  (default) or "southern"
* **"weekday"** splits variables by day of the week
* **"weekend"** splits variables by Saturday, Sunday, weekday
* **"daylight"** splits variables by nighttime/daytime. Note the user must supply a `longitude` and `latitude`
* **"dst"** splits variables by daylight saving time and non-daylight
  saving time (see manual for more details)
* **"wd"** if wind direction (`wd`) is available `type = "wd"` will split
  the data up into 8 sectors: N, NE, E, SE, S, SW, W, NW.
* **"seasonyear"** (or **"yearseason"**) will split the data into
  year-season intervals, keeping the months of a season together. For
  example, December 2010 is considered as part of winter 2011 (with
  January and February 2011). This makes it easier to consider
  contiguous seasons. In contrast, `type = "season"` will just split
  the data into four seasons regardless of the year.

If a categorical variable is present in a data frame e.g. `site` then
that variables can be used directly e.g. `type = "site"`.

## Controlling font size {#sec:font-size}

All [openair]{.pkg} plot functions have an option `fontsize`. Users can
easily vary the size of the font for each plot e.g.

```{r eval=FALSE}
polarPlot(mydata, fontsize = 20)
``` 

The font size will be reset to the default sizes once the plot is
complete. Finer control of individual font sizes is currently not
easily possible. 


## Using colours {#sec:colours}

Many of the functions described require that colour scales are used;
particularly for plots showing surfaces. It is only necessary to
consider using other colours if the user does not wish to use the
default scheme, shown at the top of \autoref{fig:colours}. The choice
of colours does seem to be a vexing issue as well as something that
depends on what one is trying to show in the first place. For this
reason, the colour schemes used in [openair]{.pkg} are very flexible: if you
don't like them, you can change them easily. R itself can handle
colours in many sophisticated ways; see for example the
`RColorBrewer` package.

Several pre-defined colour schemes are available to make it easy to
plot data. In fact, for most situations the default colour schemes
should be adequate. The choice of colours can easily be set; either by
using one of the pre-defined schemes or through a user-defined
scheme. More details can be found in the [openair]{.pkg} `openColours`
function. Some defined colours are shown in \autoref{fig:colours},
together with an example of a user defined scale that provides a
smooth transition from yellow to blue. The code that produced this
plot is shown for \autoref{fig:colours}:^[This is given for
  interest, the user does not need to know this to use the colours.]

\begin{figure}
```{r Colours,w=5,h=5,out.width='.6\\textwidth'}
library(openair)
## small function for plotting
printCols <- function(col, y) {
    rect((0:200) / 200, y, (1:201) / 200, y + 0.1, col = openColours(col, n = 201),
         border = NA)
    text(0.5, y + 0.15, deparse(substitute(col)))
}

## plot an empty plot
plot(1, xlim = c(0, 1), ylim = c(0, 1.6), type = "n", xlab = "", ylab = "",
     axes = FALSE)
printCols("default", 0)
printCols("increment", 0.2)
printCols("heat", 0.4)
printCols("jet", 0.6)
printCols("hue", 0.8)
printCols("brewer1", 1.0)
printCols("greyscale", 1.2)
printCols(c("tomato", "white", "forestgreen" ), 1.4)
```
\caption{Pre-defined colour scales in [openair]{.pkg}. The top colour
      scheme is a user-defined one.}
    {#fig:colours}
\end{figure}

The user-defined scheme is very flexible and the following provides
examples of its use. In the examples shown next, the `polarPlot}
function is used as a demonstration of their use.

```{r eval=FALSE}
# use default colours - no need to specify
polarPlot(mydata)

# use pre-defined "jet" colours
polarPlot(mydata, cols = "jet")

# define own colours going from yellow to green
polarPlot(mydata, cols = c("yellow", "green"))

# define own colours going from red to white to blue
polarPlot(mydata, cols = c("red", "white", "blue"))
```

## Automatic text formatting {#sec:quickText}

[openair]{.pkg} tries to automate the process of annotating
plots. It can be time-consuming (and tricky) to repetitively type in
text to represent $\mu$g m^-3^ or PM~10~ ($\mu$g m^-3^) etc. in R. For this reason, an
attempt is made to automatically detect strings such as `nox` or
`NOx` and format them correctly.  Where a user needs a y-axis label
such as NO~x~ ($\mu$g m^-3^) it will only be necessary to type `ylab = "nox (ug/m3)"`. The same is also true for plot titles.

Users can override this option by setting it to FALSE.


## Multiple plots on a page {#sec:multiple-plots-page}

We often get asked how to combine multiple plots on one page. Recent
changes to [openair]{.pkg} makes this a bit easier. Note that because
[openair]{.pkg} uses `lattice} graphics the base graphics `par}
settings will not work.

It is possible to arrange plots based on a column $\times$ row
layout. Let's put two plots side by side (2 columns, 1 row). First it
is necessary to assign the plots to a variable:

```{r eval=FALSE}
a <- windRose(mydata)
b <- polarPlot(mydata)
```

Now we can plot them using the split option:

```{r eval=FALSE}
print(a, split = c(1, 1, 2, 1))
print(b, split = c(2, 1, 2, 1), newpage = FALSE)
```

In the code above for the `split` option, the last two numbers give
the overall layout (2, 1) --- 2 columns, 1 row.  The first two numbers
give the column/row index for that particular plot. The last two
numbers remain constant across the series of plots being plotted.

There is one difficulty with plots that already contain sub-plots such
as `timeVariation` where it is necessary to identify the particular
plot of interest (see the `timeVariation` help for
details). However, say we want a polar plot (`b` above) and a
diurnal plot:

```{r eval=FALSE}
c <- timeVariation(mydata)
print(b, split = c(1, 1, 2, 1))
print(c, split = c(2, 1, 2, 1), subset = "hour", newpage = FALSE)
```

For more control it is possible to use the `position`
argument. `position` is a vector of 4 numbers, c(xmin, ymin, xmax,
ymax) that give the lower-left and upper-right corners of a rectangle
in which the plot is to be positioned. The coordinate system for this
rectangle is [0--1] in both the x and y directions.

As an example, consider plotting the first plot in the lower left
quadrant and the second plot in the upper right quadrant:

```{r eval=FALSE}
print(a, position = c(0, 0, 0.5, 0.5), more = TRUE)
print(b, position = c(0.5, 0.5, 1, 1))
```

The position argument gives more fine control over the plot location.



